Decoupled Fallback: A Portable Single-Pass GPU Scan

### 2025-03-01 appendix.tex
Split Appendix

\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{listings}
\usepackage{subcaption}
\usepackage{makecell}
\newcommand{\thomas}[1]{{\footnotesize\color{orange}[Thomas: #1]}}
\newcommand{\john}[1]{{\footnotesize\color{cyan}[John: #1]}}
\newcommand{\raph}[1]{{\footnotesize\color{magenta}[Raph: #1]}}

@@

\title{Decoupled Fallback: A Portable Single-Pass GPU Scan}

\author{Thomas Smith}
\affiliation{%
  \institution{Google}
  \country{USA}}

\author{Raph Levien}
\affiliation{%
  \institution{Google}
  \country{USA}
}

\author{John D. Owens}
\affiliation{%
  \institution{University of California, Davis}
  \country{USA}}
\additionalaffiliation{%
  \institution{Google}
  \country{USA}
}

\renewcommand{\shortauthors}{Smith et al.}

\begin{abstract}
  We present \emph{Decoupled Fallback}, a method that enables single-pass \emph{Chained Scans} to run on hardware without \emph{forward-progress guarantees} (FPG) while avoiding deadlock. Additionally, we introduce a tile state representation for \emph{Chained Scans} that does not rely on 64-bit atomics or memory barriers for correctness, along with a subgroup-size-agnostic intra-workgroup implementation. On FPG-lacking devices---Apple M1 Max and M3---\emph{Decoupled Fallback} achieves near-\emph{Memcpy} speeds for inclusive prefix sum in WGPU and realizes the full theoretically expected 50\% speedup over the slower \emph{Reduce-then-Scan} approach. We further demonstrate the resilience of \emph{Decoupled Fallback} against \emph{unfair} schedulers by simulating blocking at rates of up to 50\%, showing that it maintains superior performance over \emph{Reduce-then-Scan} even under extreme contention.
\end{abstract}

\begin{CCSXML}
  <ccs2012>
  <concept>
  <concept_id>10010147.10010169.10010170.10010174</concept_id>
  <concept_desc>Computing methodologies~Massively parallel algorithms</concept_desc>
  <concept_significance>500</concept_significance>
  </concept>
  <concept>
  <concept_id>10010520.10010521.10010528</concept_id>
  <concept_desc>Computer systems organization~Parallel architectures</concept_desc>
  <concept_significance>500</concept_significance>
  </concept>
  <concept>
  <concept_id>10003752.10003753.10003761.10003762</concept_id>
  <concept_desc>Theory of computation~Parallel computing models</concept_desc>
  <concept_significance>500</concept_significance>
  </concept>
  <concept>
  <concept_id>10003752.10003753.10003761</concept_id>
  <concept_desc>Theory of computation~Concurrency</concept_desc>
  <concept_significance>500</concept_significance>
  </concept>
  <concept>
  <concept_id>10010147.10010169.10010175</concept_id>
  <concept_desc>Computing methodologies~Parallel programming languages</concept_desc>
  <concept_significance>500</concept_significance>
  </concept>
  </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Computing methodologies~Massively parallel algorithms}
\ccsdesc[500]{Computer systems organization~Parallel architectures}
\ccsdesc[500]{Theory of computation~Parallel computing models}
\ccsdesc[500]{Theory of computation~Concurrency}
\ccsdesc[500]{Computing methodologies~Parallel programming languages}

\keywords{scan, parallel prefix, prefix sum, forward-progress guarantee, GPU, WGSL, decoupled fallback}

%\received{?}
%\received[revised]{?}
%\received[accepted]{?}

\maketitle

\section{Introduction}
Scan is one of the most ubiquitous operations in GPU computing, with applications in sorting~\cite{adinets2022onesweepfastersignificantdigit}, sparse matrix operations~\cite{BellGarland2009}, and more. The current state-of-the-art prefix scan algorithm, Merrill and Garland's \emph{Decoupled Lookback}, achieves near-ideal performance on NVIDIA hardware via CUDA by leveraging the \emph{Chained Scan} architecture. This approach hybridizes serial and parallel strategies to read and write each element exactly once, achieving the theoretical minimum global memory movement of $O(2n)$. \emph{Chained Scans} that fully saturate device memory bandwidth reach the \emph{speed-of-light} (SOL) performance limit---the theoretical maximum throughput constrained by the physical limits of the device. However, \emph{Decoupled Lookback} relies on architectural and language-specific guarantees provided by the NVIDIA ecosystem, most notably a device scheduler property known as a \emph{forward-progress guarantee}; without this guarantee, it risks indefinite starvation and will never complete. In practice, it is easy to demonstrate this highly undesirable behavior.

As the compute capabilities of major graphics APIs---Vulkan, Direct3D, and Metal---have grown, there has been increasing interest in adapting \emph{Decoupled Lookback} beyond CUDA for broader adoption in real-time graphics and GPU computing. However, portable implementation is constrained by both hardware and API limitations. In terms of hardware, some GPU schedulers---most notably Apple's M series and ARM---lack forward-progress guarantees, and subgroup sizes vary across architectures. In terms of API design, the absence of 64-bit atomic operations and explicit memory barriers further complicates implementation, as both are integral to \emph{Decoupled Lookback}.

This work is guided by two main goals. (1) \textbf{Portability:} We aim to develop a scan implementation that retains the benefits of the \emph{Chained Scan} architecture while being suitable for a diverse range of hardware vendors and architectures, including those without forward-progress guarantees (FPG) or support for 64-bit atomics. \textbf{Performance:} We aim for near-SOL execution to the greatest extent possible allowed by the underlying hardware and programming model.

However, certain aspects fall outside the scope of this work. \textbf{Performance guarantees} are not universally attainable, as not all architectures offer atomic operations that are sufficiently fast or scheduling models that are sufficiently fair to attain SOL performance. \textbf{Subgroup divergence} is another challenge we acknowledge but do not address directly, as shading languages do not provide explicit divergence control like CUDA's subgroup masking mechanisms. Lastly, we do not investigate alternative $O(2n)$ scan architectures beyond \emph{Chained Scan}.

We present \emph{Decoupled Fallback}, a fully portable prefix scan capable of achieving SOL performance without requiring forward-progress guarantees (FPG). Our implementation targets portability across modern GPU hardware and software platforms. It is implemented in the WebGPU Shading Language (WGSL)---an apex-level shading language that transpiles to backend API languages---ensuring seamless integration with modern graphics APIs. Our method incorporates a novel tile state representation that eliminates reliance on 64-bit atomics or memory barriers while maintaining correctness, along with a subgroup-size-agnostic intra-workgroup implementation that enables subgroup acceleration across all architectures.

\section{Background}
\subsection{Virtualization and Scheduling in the GPU programming model}
Contemporary GPUs are hierarchically organized, massively parallel processors designed to prioritize throughput over latency. As comprehensive descriptions of the GPU programming model~\cite{10.1145/1365490.1365500} already exist, we focus on the aspects most relevant to this work: scheduling and synchronization.

On GPU hardware, scheduling is divided into two levels: a \emph{workgroup} scheduler, which manages kernel launches and maps virtual processors, workgroups, to physical processors, \emph{multiprocessors}, and a \emph{subgroup} scheduler, responsible for selecting a subset of the currently \emph{occupant} subgroups on a multiprocessor for execution. The GPU programming model virtualizes processors to ensure GPU programs---\emph{kernels}---remain portable across hardware with differing physical resources. Effective virtualization requires \emph{oversubscription}, a programming paradigm where tasks are (over)partitioned in such a way that kernels request more virtual processors than are physically available. As a result, a single multiprocessor may host multiple workgroups simultaneously. However, as a consequence of virtualization, the developer relinquishes control over the order in which workgroups are launched, and once a workgroup begins execution on a multiprocessor, it must run to completion and cannot context-switch.

\subsubsection{Inter-Workgroup Barriers}
Parallel programs in general require synchronization between multiple threads (those that don't are considered embarrassingly parallel). A standard technique for synchronization is \emph{barriers.} GPUs provide a limited palette of barrier primitives. At the coarsest granularity, \emph{pipeline barriers} fully synchronize both control and memory between multiple kernel launches. However, pipeline barriers are extremely expensive and efficient algorithms strictly minimize their use. At a much finer granularity, \emph{workgroup barriers} provide reasonably efficient synchronization between the threads in a workgroup. However, synchronization between workgroups is more limited. Standard GPU programming environments provide no control barrier between workgroups.\footnote{CUDA recently introduced \emph{Thread Block Cluster} synchronization~\cite{NvidiaCudaGuide}, but it is almost certainly designed to operate within the \emph{Persistent Thread}~\cite{gupta2012} paradigm, rather than serving as a barrier across an unbounded launch.} The availability of a memory-only barrier varies across GPU environments; it is present in Direct3D and Vulkan, but was missing in versions of Metal prior to 3.2. Because of its non-portability, it is also absent from the WebGPU specification. Even when available, the use of such memory barriers causes performance issues as they are usually implemented as a full cache flush.

\subsubsection{Fairness and Progress}
Because multiple workgroups may reside on a single multiprocessor, two key issues arise at the subgroup scheduling level: \emph{fairness}---how evenly execution resources are distributed among subgroups---and \emph{progress guarantees}---ensuring that subgroups eventually make progress towards termination. However, the term ``fairness'' is used differently across the literature, leading to potential confusion. Sorensen et al.~\cite{sorensen2016,sorensen2018,sorensen2021}, whose work provides the most comprehensive taxonomy of progress models to date, use ``fairness'' specifically to describe differing levels of progress guarantees. In contrast, NVIDIA publications~\cite{4523358,Merrill2016} define ``fairness'' in terms of the distribution of execution resources among subgroups. In this work, we adopt NVIDIA's terminology.

\subsection{The Scan Primitive}
\label{sec:the-scan-primitive}
The study of \emph{scan} (\emph{parallel prefix}) networks traces back to the design of carry-lookahead adder circuits and beyond~\cite{5219801,10.5555/1098666}. A scan is typically defined on a monoid \( M \), characterized by a binary reduction operator \( \oplus \) and an identity element \( \varepsilon \). The binary operator \( \oplus \) satisfies the closure property \( \forall a, b \in M, \ (a \oplus b) \in M \) and has an identity element \( \exists \varepsilon \in M, \ \forall a \in M, \ \varepsilon \oplus a = a \). To allow parallelization, \( \oplus \) must be associative, but it is not necessarily commutative, as demonstrated in structures like the \emph{bicyclic semigroup}~\cite{Clifford1961}\footnote{Despite its name, this is a monoid.} and Rabin-Karp's \emph{second fingerprinting method for string matching}~\cite[Section 6]{Karp:1987:ERP}. In a scan, the result at the \( n \)-th element is the reduction of the preceding subsequence of elements. If the subsequence includes the \( n \)-th element, it is called \emph{inclusive}; if it excludes the \( n \)-th element, it is called \emph{exclusive}. The most common scan type is the prefix sum, where \( \oplus \) is addition. For example:
\[
  x = [x_1, x_2, x_3, \dots, x_n] \ \ \ \ \ \ \ y = [1, 1, 1, 1, 1]
\]
\[
  \text{InclusiveScan}(x, \oplus) = [x_1, x_1 \oplus x_2, x_1 \oplus x_2 \oplus x_3, \dots, x_1 \oplus x_2 \oplus \cdots \oplus x_n]
\]
\[
  \text{InclusiveScan}(y, +) = [1, 2, 3, 4, 5]
\]
Due to its significance in circuits and as a fundamental algorithmic primitive, scan has been extensively studied in both electrical engineering and computer science. Harris~\cite{1292373} offers a taxonomy that relates depth, fanout, and wire tracks, while Hinze~\cite{10.1007/978-3-540-27764-4_11} develops an algebraic framework for scans. Snir~\cite{10.1016/0196-67748690003-9} proved that depth $d$ and size $s$ are related by $s + d \ge 2n - 2$, and Fich~\cite{10.1145/800061.808738} proved that among minimum depth scans Ladner-Fischer~\cite{10.1145/322217.322232} networks have optimal size. Blelloch~\cite{Blelloch:1989:SAP} adapted scan to the PRAM model and popularized its use as an algorithmic primitive. Finally, Merrill and Garland~\cite{Merrill2016} provide a review of GPU scan implementations, complementing earlier work by Merrill and Grimshaw~\cite{Merrill2009}.

\subsection{Evolution of Inter-Workgroup Scan Architectures}
Contemporary GPU scans depart from early PRAM-like scans~\cite{Horn, Hensley, Sengupta2006, Gress} by leveraging the GPU memory hierarchy, which consists of progressively faster but increasingly private tiers. Efficient scans integrate fine-grained intra-workgroup (\emph{local}) strategies in shared memory and registers with coarse-grained inter-workgroup (\emph{global}) strategies in global memory. Given their low arithmetic intensity, local scans are inherently memory-bound, as demonstrated by Merrill and Grimshaw~\cite{Merrill2009}. Consequently, recent scan optimizations have shifted toward refining global strategies, particularly inter-workgroup coordination and synchronization. This section reviews three key approaches---\emph{Scan-then-Propagate}, \emph{Reduce-then-Scan}, and \emph{Chained Scan}---analyzing their handling of inter-workgroup dependencies and efficiency in minimizing global memory access while maximizing bandwidth utilization.

\subsubsection{Scan-then-Propagate}
Introduced by Sengupta et al.~\cite{10.5555/1280094.1280110}, \emph{Scan-then-Propagate}~\cite{GPUGems3, Sengupta2011} extends intra-workgroup scans to large inputs by resolving inter-workgroup dependencies in three phases:
\begin{enumerate}
  \item \textbf{Intra-Workgroup (Local) Scan:}  Each workgroup performs an inclusive scan on its assigned work tile, storing local results and reductions in global memory.
  \item \textbf{Spine Scan:} The workgroup reductions, collectively forming the \emph{spine}, are gathered and scanned to compute the \emph{root}, resolving dependencies between work tiles. For large scans, this stage is an insignificant amount of work compared to the first and third.
  \item \textbf{Propagation:} Each workgroup retrieves its portion of the root and adds it to the local scan output.
\end{enumerate}
The absence of inter-workgroup barriers necessitates separate kernel launches and pipeline barriers for dependency resolution, naturally enforcing this structure. To accommodate arbitrarily large inputs, Sengupta et al.\ employ recursion, repeatedly applying the three-phase process until the spine fits within a single work tile. Thus, an input of size $n$ and a work tile of size $t$ results in a recursive depth of $\lceil \log_t n \rceil$, and $2\cdot\lceil \log_t n \rceil - 1$ kernel launches. Each recursive step $k$ processes $n/t^k$ tiles, leading to a total of $\sum_{k=1}^{\lceil \log_t n \rceil} n/t^k \approx (n - 1)/(t - 1)$ work tiles over all steps. Since each tile moves $4t$ data ($2t$ local scan, $2t$ propagation), the total global data movement is $O\left(\frac{4t(n - 1)}{t - 1}\right) = O(4n)$.

\subsubsection{Reduce-then-Scan}
If memory bandwidth resources are sufficiently scarce relative to compute, memory-bound algorithms like scan can profit by trading additional computation for reduced memory traffic. This trade-off led to a refinement of the \emph{Scan-then-Propagate} strategy called \emph{Reduce-then-Scan}~\cite{10.1145/1375527.1375559, Merrill2009, 10.1109/TPDS.2012.336, 10.5555/2031978.2032029}: by storing only the work tile reduction in global memory during the first phase and performing a redundant scan in the third, it eliminates $n$ global data movement while preserving the three-phase structure:
\begin{enumerate}
  \item \textbf{Work Tile Reduction:} Each workgroup performs a pure reduction on its work tile partition, posting the result to global memory.
  \item \textbf{Spine Scan:} The spine is scanned to produce the root.
  \item \textbf{Intra-Workgroup Scan and Propagation:} Each workgroup performs the local scan on the work tile, incorporating its portion of the root as it writes to global memory.
\end{enumerate}

Merrill and Grimshaw~\cite{Merrill2009} introduced \emph{workgroup raking} to reduce recursion overhead for arbitrarily large inputs. Instead of recursively reducing the spine, the input is partitioned among a fixed number of workgroups, $c$, where $c \geq o$, the total workgroup occupancy. This constrains the spine from a variable, potentially recursive structure to a fixed size $c$, ensuring a recursion depth of at most two. As a result, global memory movement for the spine scan is reduced from $O(n/t)$ to $O(2c)$, and kernel launches and pipeline barriers are limited to three, yielding a total global memory movement of $O(3n + 4c) = O(3n)$.

Breitbart~\cite{10.5555/2031978.2032029} applied Gupta et al.'s \emph{Persistent-Threads}~\cite{gupta2012} to scan, eliminating kernel launch overheads. Recognizing that raking naturally aligns with this paradigm, Breitbart's method statically \emph{discovers} the workgroup occupancy and launches exactly $o$ kernels. Since multiprocessors are not oversubscribed, workgroups persist for the entire kernel duration, enabling synchronization at an atomics-based inter-workgroup memory and control barrier without kernel launches or pipeline barriers. While global memory movement remains $O(3n)$, this consolidates the scan into a single kernel, minimizing launch and pipeline overheads.

\subsubsection{Chained Scan, StreamScan}
First introduced by Yan et al.~\cite{10.1145/2442516.2442539} in \emph{StreamScan}, the key innovation of the \emph{Chained Scan} architecture lies in its hybridization of parallel and serial strategies: achieving parallelism at the intra-workgroup level while minimizing global data movement through serial scan operations at the inter-workgroup level. Instead of using kernel launches and pipeline barriers as inter-workgroup synchronization points, \emph{Chained Scans} launch a single kernel that assigns work tiles in a serial order to workgroups, utilizing atomics and bit-packed status flags to ensure coherent views of data. To enforce this serial ordering, work tiles are dynamically assigned using atomic increment operations, rather than arbitrary assignment based on virtualized workgroup indices. This shifts dependency resolution from explicit synchronization via barriers to implicit synchronization via access order. As a result, each element is read and written exactly once, achieving the theoretical minimum of $O(2n)$.

In any \emph{Chained Scan}, once a workgroup computes its local scan, it must complete two objectives: communicate its reduction to successor tiles and resolve dependencies by incorporating the reductions of all preceding tiles. This process, which we refer to as \emph{Chained Propagation}, defines the set of inter-workgroup techniques that facilitate dependency resolution. In \emph{StreamScan}, these two objectives are combined into a single phase, where each workgroup waits for its immediate predecessor to complete, ingests the predecessor's reduction, and posts the resulting \emph{inclusive} reduction for its immediate successor to consume. This approach depends on a forward-progress guarantee of the GPU's execution model. Given such a guarantee, the predecessor will eventually produce a result and broadcast it. On devices lacking such a guarantee, the workgroup may wait indefinitely, resulting in deadlock.

While \emph{StreamScan} successfully reduces per-element device memory accesses to two, it does not consistently saturate global memory bandwidth, preventing it from achieving SOL performance. This limitation arises because a workgroup in \emph{StreamScan} cannot post its reduction to global memory until it has received the result of its immediate predecessor. While this results in efficient $O(n/t)$ communication of reductions, and $O(2n+ n/t)= O(2n)$ total data movement, the sequential dependency creates a bottleneck that limits performance to the rate at which reductions propagate through device memory. Although this can be mitigated by increasing the size of the work tile, limited on-chip resources preclude scaling the tile size indefinitely. Moreover, this approach is unsuitable from a portability perspective, as excessively large work tiles risk reducing occupancy on less capable hardware.

\subsubsection{Chained Scan, Decoupled Lookback}
Introduced by Merrill and Garland, the \emph{Decoupled Lookback} technique refines \emph{Chained Propagation} by separating reduction communication from dependency resolution, mitigating bottlenecks caused by propagation latency. In this approach, each workgroup posts its local reduction to global memory and then performs a \emph{lookback}---a serial traversal backward along the spine---to compute its inclusive reduction. Work and memory accesses are constant-bounded to the workgroup occupancy $o$ by posting the inclusive reduction after the lookback. However, granting each workgroup control over its dependency resolution does not eliminate reliance on \emph{progress guarantees}, as workgroups must still spin on tiles that have yet to post their reductions, risking deadlock.

Although \emph{Decoupled Lookback} increases global memory access during the propagation phase to $O(o(n/t))$, the serial ordering of work tile processing confines spine memory accesses to a roughly $o$-width sliding window, making them highly likely to be cache-resident. As a result, the additional global memory traffic along the spine introduces negligible overhead. With total global memory movement of $O(2n+o(n/t))= O(2n)$, \emph{Decoupled Lookback} achieves full utilization of device memory bandwidth and delivers SOL performance without the work tile size tuning required by \emph{StreamScan}.

\subsection{Ideal Scan Architecture}
\emph{Chained Scan} and \emph{Reduce-then-Scan} represent two contrasting paradigms in scan design. By eliding the barrier-idiom, \emph{Chained Scan} aligns with the oversubscription paradigm, delegating occupancy management to the workgroup scheduler. In contrast, \emph{Reduce-then-Scan} employs workgroup raking, requiring fixed launches and explicit occupancy control, forcing developers to manually tune launch sizes or implement occupancy discovery techniques.

Hardware-level considerations underscore the limitations of the fixed-launch approach. On GPUs where the global memory data cache is indexed by physical addresses---and therefore requires lookups in \emph{translation lookaside buffers} (TLBs)---fixed launches can lead to performance degradation due to cache thrashing. This issue occurs when task sizes are large enough that each workgroup accesses its own unique page entry, and the device lacks sufficient TLB coverage to support the memory access patterns of all workgroups. Although this behavior typically does not manifest with simple monoids, more complex monoids with \emph{struct-of-array} memory access patterns---or applications like radix sorting, where the scattering phase of each pass results in $radixDigits \cdot o$ unique memory locations---may experience significant performance degradation.

Given these constraints, oversubscription and \emph{Chained Scan} emerge as the more portable and preferred architecture. Furthermore, its single-kernel design and serial ordering make it suitable for in-situ operations, benefiting applications such as data deduplication, sparse data compaction, error correction, and interval merging. However, it has a critical limitation: its reliance on a forward-progress guarantee. While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that ARM and Apple's GPUs do not. Without forward-progress guarantees, \emph{Chained Scan} is not merely inefficient---it risks indefinite deadlock, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.

\subsection{Progress Guarantees and Their Impact on \emph{Chained Scan}}
A \emph{forward-progress guarantee} ensures that all occupant subgroups---of possibly different workgroups---eventually make progress toward termination, preventing indefinite starvation. Prior to Sorensen et al.~\cite{sorensen2018,sorensen2021}, developers largely relied on untested, folklore-like assumptions about GPU progress guarantees. Instead, Sorensen et al.\ systematically proposed and tested three progress models:\footnote{Note, a ``workgroup'' in Sorensen et al.'s definition is not necessarily in-flight and occupant on a multiprocessor.}
\begin{itemize}
  \item \textbf{Occupancy-Bound Execution (OBE)}: Workgroups that have made at least one execution step.
  \item \textbf{Heterogeneous System Architecture~\cite[Section~2.11]{HSASysArch12} (HSA)}: The workgroup with the lowest ID that has not yet terminated.
  \item \textbf{Linear OBE (LOBE)}: Workgroups that have made at least one step, and any workgroup with a lower ID than a workgroup that has made at least one step.
\end{itemize}
Their 2021 findings~\cite{sorensen2021} demonstrated that Apple and ARM GPUs do not support LOBE, contradicting their prior hypothesis~\cite{sorensen2018}.

Sorensen et al.\ classify \emph{Chained Scan} as a \emph{one-way producer-consumer} (PC) HSA algorithm: a workgroup only consumes from lower-ID counterparts, ensuring that execution proceeds in a strictly ordered fashion. HSA guarantees termination, as the lowest ID workgroup is always given forward progress. If subgroups from the lowest ID workgroup are not guaranteed execution, then dependent workgroups may spin indefinitely, waiting on an update that never arrives.

However, the HSA classification relies on the assumption that \emph{Chained Scan} follows scheduler-assigned virtualized IDs rather than dynamically acquired atomic-bumped IDs, as used by both \emph{StreamScan} and \emph{Decoupled Lookback}. Since virtual IDs do not necessarily match atomic IDs, HSA does not guarantee termination---the lowest virtual ID may not be scheduled first, potentially leading to starvation.

LOBE adds an additional condition on top of OBE\@: lower-ID workgroups must be scheduled before higher-ID ones. However, this condition is unnecessary for \emph{Chained Scan} for two reasons: (1) We already guarantee serial ordering of work items through atomic bumping, making OBE sufficient. (2) \emph{Chained Scan} only requires forward progress after a workgroup has atomically acquired its tile index. By definition, a workgroup that has acquired its index must be occupant on a physical processor (i.e., it has made at least one execution step); thus, OBE guarantees forward progress.

Thus, by dynamically acquiring IDs through atomics, \emph{Chained Scan} shifts its progress model requirement from HSA to OBE. While LOBE encompasses both HSA and OBE, its stricter ordering guarantees are unnecessary for correctness. Ultimately, regardless of a device's specific progress model, any architecture lacking forward-progress guarantees will fail to execute \emph{Chained Scan}. In short:

\begin{table}[h!]
  \small
  \centering
  \begin{tabular}{lcccc}
    \toprule
    Algorithm                             & LOBE & OBE & HSA & None \\
    \midrule
    \emph{Chained Scan} with Workgroup ID & Yes  & No  & Yes & No   \\
    \emph{Chained Scan} with Atomic ID    & Yes  & Yes & No  & No   \\
    Ours                                  & Yes  & Yes & Yes & Yes  \\
    \bottomrule
  \end{tabular}
  \caption{Comparison of \emph{Chained Scan} Variants and Our Implementation Across Progress Models.\label{tab:progress_models}}
\end{table}

\subsection{Why is Reliance on Forward-Progress Guarantees a Portability Problem?}
The absence of FPG in key hardware segments poses a critical challenge for developers. Our testing (Figure~\ref{fig:dist}) shows that, at best, running \emph{Chained Scan} without FPG results in performance worse than \emph{Reduce-then-Scan}, and frequently triggers \emph{Timeout Device Recovery}\label{sec:tdr}\footnote{Operating systems handle unresponsive programs differently. On Windows, the Timeout Detection and Recovery (TDR) system resets the GPU after approximately 2 seconds. Other operating systems---such as macOS and Linux---employ their own timeout mechanisms with varying behaviors. For consistency, we adopt Windows’ 2-second limit as a reasonable expectation for maximum execution time.} (TDR), crashing the program.
\begin{figure}
  \centering
  \begin{subfigure}[b]{0.45\linewidth}
    \includegraphics[width=\linewidth]{graphics/decoupled_lookback.pdf}
    \caption{\emph{Decoupled Lookback.}}
    \label{fig:decoupled}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.45\linewidth}
    \includegraphics[width=\linewidth]{graphics/ours.pdf}
    \caption{\emph{Decoupled Fallback (ours).}}
    \label{fig:ours}
  \end{subfigure}
  \caption{Histogram plots, 500 runs of Inclusive Prefix Sum, $\mathbf{2^{20}}$ 32-bit elements, on an FPG-lacking device, Apple M1 Max. Bar heights ($y$ axis) indicate the number of runs that completed in the execution time indicated on the $x$ axis. The TDR threshold is Windows's threshold for resetting the GPU\@; as noted in Section~\ref{sec:tdr}, MacOS handles this differently.}
  \label{fig:dist}
\end{figure}
Exacerbating this issue, no major graphics API provides a mechanism to query FPG support. This lack of transparency forces developers to design algorithms without knowing whether critical guarantees are available on the target hardware. In shader environments, where cross-platform portability is essential, this uncertainty compels a fallback to older, less efficient strategies like \emph{Reduce-then-Scan}, sacrificing performance for broader compatibility.

\subsection{Earlier Attempts at Portability}
Our first attempt \raph{we'll add the cite post-review} at implementing a \emph{Decoupled Lookback} scan without relying on FPG was \emph{Scalar Fallback}. Each time a workgroup spins waiting on the reduction from a preceding tile, it advances a \emph{fallback} operation by processing one scalar element from that tile. This eliminates the possibility of deadlock because the number of such operations is bounded; the wait will eventually terminate, possibly at the expense of doing redundant work.

\emph{Scalar Fallback} suffers from two primary inefficiencies. First, the fallback reduction is performed serially by a single thread, progressing only when the lookback thread performs a waiting spin. This results in a $t$-sized blocking tile requiring $t$ spins to complete. Second, once the fallback completes, the lookback thread does not post its result to device memory, forcing every blocked workgroup to redundantly recompute the fallback. While \emph{Scalar Fallback} achieved SOL performance on hardware with FPG and executed correctly without it, its performance on the latter was inferior to \emph{Reduce-then-Scan}.

\section{Decoupled Fallback}
\emph{Decoupled Fallback} addresses both inefficiencies. Our key advance is to utilize the parallelism available in the workgroup to perform the fallback computation, rather than performing it sequentially. To do so requires nontrivial coordination within the workgroup, given that the \emph{Chained Propagation} logic is typically performed by the first thread.

\begin{algorithm}
  \small
  \SetAlgoLined%
  \KwGlobalIn{%
    $\textit{in}[]$,
    $\textit{out}[]$,
    $\textit{spine}[]$,
    $\textit{tileBump}$,
  }
  \KwSharedIn{%
    $\textit{wg\_partials}[]$,
    $\textit{wg\_fallback}[]$,
    $\textit{wg\_prev\_reduction}$,
    $\textit{wg\_broadcast}$,
    $\textit{wg\_control}$,
  }
  \KwGlobalOut{$\textit{out}[]$ inclusive scan of $\textit{in}[]$}
  \KwConstants{Vectors per thread $\textit{p}$}

  \If{$\textit{threadid} = 0$}{
    $\textit{wg\_broadcast} \gets \textnormal{atomicAdd}(\&\textit{tileBump}, 1)$\;
    $\textit{wg\_control} \gets \textnormal{LOCKED}$\;
  }
  \textbf{workgroupBarrier()}\;

  $\textit{tile\_id} \gets \textit{wg\_broadcast}$\;

  $\textit{local\_vectors} \gets \textnormal{array}\langle \textnormal{vec4}\langle u32\rangle, \textit{p} \rangle$\;

  \textbf{Load}($\textit{in}$, $\textit{tile\_id}$, \textit{local\_vectors})\;
  \textbf{SubgroupRake}($\textit{local\_vectors}$, $\textit{wg\_partials}$)\;
  \textbf{workgroupBarrier()}\;

  \textbf{WorkgroupWideScan}($\textit{wg\_partials}$)\;
  \textbf{workgroupBarrier()}\;

  \tcp{Chained Propagation, Decoupled Fallback}
  \textbf{PostLocalReduction}($\textit{spine}$, $\textit{tile\_id}$, $\textit{wg\_partials}$)\;
  \While{$\textit{wg\_control} = \textnormal{LOCKED}$}{
    \textbf{Lookback}($\textit{spine}$, $\textit{tile\_id}$, $\textit{wg\_prev\_reduction}$, $\textit{wg\_control}$)\;
    \textbf{workgroupBarrier()}\;

    \If{$\textit{wg\_control} = \textnormal{LOCKED}$}{
      \textbf{FallbackReduce}($\textit{in}$, $\textit{wg\_fallback}$)\;
      \textbf{FallbackInsertionAttempt}($\textit{spine}$, $\textit{wg\_fallback}$, $\textit{wg\_prev\_reduction}$, $\textit{wg\_control}$)\;
      \textbf{workgroupBarrier()}\;
    }
  }

  \textbf{Write}($\textit{out}$, $\textit{tile\_id}$, $\textit{local\_vectors}$, $\textit{wg\_partials}$, $\textit{wg\_prev\_reduction}$)\;

  \caption{High-Level Scan Kernel}
\end{algorithm}

To optimize the fallback operation, we employ three strategies. (1) First, a shared memory \emph{control flag} enables threads performing the lookback to signal the entire workgroup when a fallback is necessary. (2) Next, we limit the number of spins a thread can perform while waiting on a tile; exceeding this threshold triggers a fallback. While this approach risks \emph{false positive} fallbacks---since a late tile is indistinguishable from a genuinely blocking one---the maximum spin count can be tuned to minimize redundant fallbacks. Increasing the spin count lowers false positives but delays progress in true-blocking scenarios. False positives not only increase memory traffic but also consume compute resources that could be scheduled on the same multiprocessor. (3) After completing a fallback, the workgroup attempts to post the reduction result of the blocking tile, preventing redundant fallbacks by subsequent workgroups.

>>>>

\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{listings}
\usepackage{subcaption}
\usepackage{makecell}

@@

\appendix
\section{Artifact Notes}
\subsection{Baseline Comparisons: \emph{Memcpy}}%
\label{sec:memcpy}
\begin{table}
  \small
  \centering
  \begin{tabular}{l r r}
    \toprule
    Device                                & \makecell{Memory           \\ Bandwidth \\ (GB/s)} & \makecell{Expected \\ Throughput \\ (32-bit ele/sec)} \\
    \midrule
    Intel HD620 Single-Channel            & 17.1             & 2.13e9  \\
    Intel HD620 Dual-Channel              & 34.1             & 4.26e9  \\
    ARM Mali-G78 MP20 / Tensor (Pixel 6a) & 51.2             & 6.40e9  \\
    Apple M3 (10 Cores)                   & 102.4            & 12.8e9  \\
    Apple M1 Max (32 Cores)               & 409.6            & 51.2e9  \\
    Nvidia 2080 Super                     & 496              & 62.0e9  \\
    AMD 7900 XT                           & 800              & 100.0e9 \\
    \bottomrule
  \end{tabular}
  \caption{Vendor-Specified Memory Bandwidth and Expected Throughput (32-bit Elements per Second).\label{tab:memory_bandwidth}}
\end{table}
Although all of Dawn's backend APIs support \emph{timestamp queries inside encoders}, Apple GPUs do not, so we are forced to implement our own \emph{Memcpy} kernel to measure memory bandwidth, as opposed to simply timing an API's copy buffer operation. It is well known that \emph{observed} peak memory bandwidth tends to diverge from vendor-listed peak bandwidth, so for completeness we list these figures above. We make a number of observations:
\begin{itemize}
  \item \textbf{Intel HD620:} As our HD620 laptop is only equipped with a single memory DIMM, its peak bandwidth is limited to the lower bound, 2.13e9 ele/sec. Thus, our observed bandwidth of 1.498e9 is reasonable.
  \item \textbf{ARM Mali-G78 MP20 / Pixel 6a (Tensor):} The observed throughput is significantly lower than expected, despite a vendor-listed bandwidth of 6.40e9 ele/sec. The cause of this discrepancy remains unclear, but we note other developers have observed similar speeds.
  \item \textbf{Apple M3:} Observed throughput matches expectations.
  \item \textbf{Apple M1 Max:} Observed throughput matches expectations.
  \item \textbf{Nvidia 2080 Super:} Observed throughput matches expectations.
  \item \textbf{AMD 7900 XT:} Our above result, where \emph{Decoupled Fallback} was observed to be faster than \emph{Memcpy}, is likely due to compute latency counterintuitively improving memory access patterns, leading to reduced paging. Furthermore, we see that our \emph{Decoupled Fallback} throughput is within spec and reasonable.
\end{itemize}

\subsection{Baseline Comparisons: \emph{Reduce-then-Scan}}
\label{sec:rts}
As we desire the most competitive baseline possible, our \emph{Reduce-then-Scan} kernels employ the same intra-workgroup scan strategy and tuning parameters as \emph{Decoupled Fallback}. Initially, we adopted Merrill and Grimshaw's~\cite{Merrill2009} workgroup raking approach for inter-workgroup processing. However, empirical testing revealed that raking was consistently outperformed---by up to $\sim$15\% on lower-end devices---by a simpler approach that processes the spine serially with a single workgroup. We confirm the efficiency of our \emph{Reduce-then-Scan} implementation by observing that its performance closely aligns with the theoretically expected $O(3n)$ global memory throughput across all tested devices.
